---
title: "Lab 2: Why Start With Indicators?"
author: "Matt Harris & Michael Fichman"
date: '2024-09-05'
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Learning Goals for Today

In this lab we will:

1.  Look at examples of "wide" and "long" data.

2.  Use {tidycensus} to gather ACS data

3.  Use the {sf} package for geospatial operations such as **union**, **intersection** and **buffer**

4.  Use maps and tables to analyze change over time using the products of the geospatial operations
    
This lab (and this week's Github folder) is the "pantry" with all the ingredients for your second assignment, where you will adapt this analysis to analyze willingness to pay for transit in a separate city. Your job is to take this code base, and feed new census data and transit stops to it.

## Why Start With Indicators?

from Steif (2021), [Section
1.1](https://urbanspatial.github.io/PublicPolicyAnalytics/TOD.html#why-start-with-indicators):

This is an indicator - a stylized fact that gives simple insight into a complicated phenomena. Mastering indicators is critical for conveying nuanced context to non-technical audiences. Here are four suggestions on what makes a good indicator:

-   A *relatable* indicator is typically motivated by a pressing policy
    concern. "How is it possible that passenger rail in New York City
    has such widespread delays, service suspensions, and rider
    discontent?" A great indicator solicits interest from an audience.

-   A *simple* indicator may be used as an exploratory tool in place of
    more complex statistics. Simplicity helps the audience understand
    the indicator's significance and keeps them engaged in the analysis.

-   A *relative* indicator draws a contrast. "How can New York City
    passenger rail, with the most trips, still loose more money than
    each of the next ten largest cities?" Contextualizing an indicator
    with a relevant comparison makes for greater impact.

-   A *good* indicator typically generates more questions than answers.
    Thus, a good indicator fits into a broader narrative which helps
    motivate a more robust research agenda and ultimately, more applied
    analytics.

### Load packages and functions

We begin by loading packages, loading our API key etc.,

There are a few new wrinkles here - we use the `source` command to bring in customized functions from our text book, we set some `options` for scientific notation (`scipen`), and some `options` for `tigris` data to be imported in `sf` form (you might find that useful in your assignment). We also specify a 5 color palette called `palette5` - you might want to make your own for your work if you like different colors.

```{r setup_packages, warning = FALSE, message = FALSE}
# Load Libraries

library(tidyverse)
library(tidycensus)
library(sf)
library(kableExtra)

options(scipen=999)
options(tigris_class = "sf")

source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")

palette5 <- c("#f0f9e8","#bae4bc","#7bccc4","#43a2ca","#0868ac")
```


```{r load_key, warning = FALSE, eval = FALSE}
census_api_key("d9ebfd04caa0138647fbacd94c657cdecbf705e9", overwrite = TRUE)
```

### Use `get_acs()` to get 2016 ACS data

Notice this returns "long" data - let's examine it using `glimpse`.

What's this CRS? Check out [spatialreference.com](spatialreference.com) and look it up!

```{r results='hide'}
tracts16 <-  
  get_acs(geography = "tract",
          variables = c("B25026_001E","B02001_002E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E", "B25058_001E",
                        "B06012_002E"), 
          year=2016, state=42,
          county=101, geometry=TRUE) %>% 
  st_transform('ESRI:102728')
```

#### Wide data vs long data (and `spread`vs `gather`)

Take a moment to think about the *shape* of data. The gif below is a visualization of **wide** vs. **long** data. The gif shows the same data transformed into two different shapes. You will do this often in when working with data. **Wide** data is a more common representation and is good for all forms of data analysis, including spatial data. In this data the `id` is not repeated in the table. On the other hand, **long** data is takes all the wide-data columns and makes them rows by repeating the `id` for each piece of data. The **long** format is often good for plotting and analytics over groups within the data; not as great with spatial data.

*Exercise - stare hypnotically at this viz for a bit*

![wide vs long
data](https://github.com/mafichman/musa_5080_2024/blob/main/Week_2/images/tidyr-spread-gather.gif?raw=true)

Image: Garrick Buie

By default, `get_acs()` returns **long** data where the `GEOID` is repeated for each of the acs `variables` requested. The `output` argument in `get_acs()` can be used to return data in a **wide** format. Or we can do that ourselves with the `spread()` and `gather()` functions.


*Why do you think this data type might be useful? Can you name a couple of examples?*

### Using ggplot to map census data with {sf} and quantile symbologies

*Exercise: Let's make a few plots by filtering `B25026_001` within a ggplot statement and visualizing it. Each plot adds more and more nuance and information. Run the code, and examine each to see what we've added each time.*

We include a couple of custom functions from our textbook:

*What does `q5` do?*

*What does `qBr` do?*

(Note: the `qBr()` function used here is an update to the one used version used in the text book so the results may appear different.)

Plot 1:

```{r}
ggplot() +
  geom_sf(data = tracts16 %>% 
            filter(variable == "B25026_001"), 
          aes(fill = estimate)) +
  theme(
    plot.title = element_text(size=22)
    )
```

Plot 2:

```{r}
  ggplot() +
  geom_sf(data = tracts16 %>% 
            filter(variable == "B25026_001"),
          aes(fill = q5(estimate))) +
  theme(plot.title = element_text(size=22)) 
```

Plot 3:

```{r}
ggplot() +
  geom_sf(data = tracts16 %>% 
            filter(variable == "B25026_001"),
          aes(fill = q5(estimate))) +
  scale_fill_manual(values = palette5,
                    labels = qBr(tracts16 %>% 
            filter(variable == "B25026_001"), "estimate"),
                    name = "Total\nPopluation\n(Quintile Breaks)") +
  theme(plot.title = element_text(size=22))
```

Plot 4:

```{r}
ggplot() +
  geom_sf(data = tracts16 %>% 
            filter(variable == "B25026_001"), 
          aes(fill = q5(estimate))) +
  scale_fill_manual(values = palette5,
                    labels = qBr(tracts16 %>% 
            filter(variable == "B25026_001"), "estimate"),
                    name = "Population\n(Quintile Breaks)") +
  labs(title = "Total Population", subtitle = "Philadelphia; 2016") +
  mapTheme() + 
  theme(plot.title = element_text(size=22))
```

### Transforming Long Data to Wide Data using spread

Here we will modify the 2016 ACS by using `spread()` to widen the data  - each census variable is now a column. This looks like the data we had in Lab 1.

We will use `rename()` to give sensible names to the columns and we will use `mutate()` to make new features out of the existing columns. These are all `dplyr` "verbs" we saw in Lab 1, except here we are doing it on {sf} spatial data frames.

```{r}
tracts16 <- 
  tracts16 %>%
  dplyr::select( -NAME, -moe) %>%
  spread(key = variable, value = estimate) %>%
  rename(TotalPop = B25026_001, 
         Whites = B02001_002,
         FemaleBachelors = B15001_050, 
         MaleBachelors = B15001_009,
         MedHHInc = B19013_001, 
         MedRent = B25058_001,
         TotalPoverty = B06012_002)
```

Let's create new rate variables using mutate..

```{r}
tracts16 <- 
  tracts16 %>%
  mutate(pctWhite = ifelse(TotalPop > 0, Whites / TotalPop, 0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop), 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         year = "2016") %>%
  dplyr::select(-Whites,-FemaleBachelors,-MaleBachelors,-TotalPoverty)
```

Tracts 2016 is now complete. Let's grab 2020 tracts in "wide" format and do the same renaming.

### 2020 Census Data

Notice that we are getting "wide" data here in the first place. This saves us the trouble of using `spread`. We do this by using the `output="wide"` argument to `get_acs()`. In previous code chunks we split the use of `get_acs()` to download the data from the use of `mutate()` and other {dplyr} functions. Here we put the entire process into one continuous sequence of code using the "pipe" operator `%>%`.

```{r results='hide'}
tracts20 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E"), 
          year=2020, 
          state=42, 
          county=101, 
          geometry=TRUE, 
          output="wide") %>%
  st_transform('ESRI:102728') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         year = "2020") %>%
  dplyr::select(-Whites, -FemaleBachelors, -MaleBachelors, -TotalPoverty) 
```

To combine the 2016 and 2020 data into the same data frame we use `rbind()`. The "r" in `rbind()` stands for "row". So this function "binds rows of two or more data frames". Similarity the `cbind()` function binds columns of data frames, but that function is not used in this lab.

*What is the value of combining data in this structural form??*

```{r}

allTracts <- rbind(tracts16,tracts20)

```

### Wrangling Transit Open Data

We want to relate transit stops to our tracts, so let's bring in Philadelphia's two main subway lines as spatial data. To make a spatial data frame of the whole system, we `rbind` them together into one object, where each row is a stop, and the columns represent the line and the name of the stop. 

*Run this code on your own - look at the objects `el` and `Broad_St` individually before you `rbind` them. Think about why the `rbind` is structured this way - you might have to think about how you create your transit data set in your homework.*

We select only the `Station` and `Line` information (see how we do this in the rbind!), and then `st_transform` the whole thing to be the same crs as `tracts16`

```{r}

el <- st_read("https://opendata.arcgis.com/datasets/8c6e2575c8ad46eb887e6bb35825e1a6_0.geojson")
Broad_St <- st_read("https://opendata.arcgis.com/datasets/2e9037fd5bef406488ffe5bb67d21312_0.geojson")

septaStops <- 
  rbind(
     el %>% 
      mutate(Line = "El") %>%
      dplyr::select(Station, Line),
     Broad_St %>%
      mutate(Line ="Broad_St") %>%
      dplyr::select(Station, Line)) %>%
  st_transform(st_crs(tracts16))  
```

Let's visualize it

```{r}

ggplot() + 
  geom_sf(data=st_union(tracts16)) +
  geom_sf(data=septaStops, 
          aes(colour = Line), 
          show.legend = "point", size= 2) +
  scale_colour_manual(values = c("orange","blue")) +
  labs(title="Septa Stops", 
       subtitle="Philadelphia, PA", 
       caption="Figure 2.5") +
  mapTheme()
```

### Relating SEPTA Stops and Tracts

Create buffers (in feet - note the CRS) around Septa stops - Both a buffer for each stop, and a union of the buffers... and bind these objects together.

Let's do this in pieces to understand this hefty code chunk.

We put all these buffers in the same data frame... this will come in handy in your assignment when we need info about not just TOD/non-TOD indicators, but indicators by stop... 

*What data format are these - wide or long? Check out the maps we make afterwards using facet_wrap - this requires long data.*

```{r}

stopBuffer <- st_buffer(septaStops, 2640)

stopUnion <- st_union(st_buffer(septaStops, 2640))

septaBuffers <- 
  rbind(
     stopBuffer %>%
      mutate(Legend = "Buffer") %>%
      dplyr::select(Legend),
     stopUnion %>%
      st_sf() %>%
      mutate(Legend = "Unioned Buffer"))
```

Let's examine both buffers by making a "[small multiples](https://www.juiceanalytics.com/writing/better-know-visualization-small-multiples)" plot. This type of plot shows multiple views of the same data or geographic area. We accomplish this by using the `facet_wrap()` {ggplot2} function on the *Legend* column.

```{r}
ggplot() +
  geom_sf(data=septaBuffers) +
  geom_sf(data=septaStops, show.legend = "point") +
  facet_wrap(~Legend) + 
  labs(caption = "Figure 2.6") +
  mapTheme()
```

### Using the `sf` Package for Spatial operations

Consult the text to understand the difference between these three types of joins and discuss which is likely appropriate for this analysis

Create an sf object with ONLY the unioned buffer from the `septaBuffers` spatial data frame.

```{r}
buffer <- filter(septaBuffers, Legend=="Unioned Buffer")
```

### Spatial Intersection of tracts and buffers

In this section, we deep dive a bit into three different ways to achieve a similar output. Namely, what we are accomplishing is to select the tracts from `tracts16` that intersect the buffer of transit stop in `buffer` . The three different ways to think about this operation are:

-   Clip - the intersection returns only the areas where both `buffer` and
    `tracts16` polygons overlap. This is often referred to as "clipping"

-   Intersect - The intersection polygons are returns the full extent of any polygon
    in `tracts16` that touch (or intersect) the `buffer`. There is no
    clipping of the polygons

    -   We will see three different ways to accomplish this particular
        outcome

-   Centroid join - The intersection of polygons to centroids returns the full extent of
    any polygon in `tracts16` that has its center (called a centroid)
    within the `buffer`

The version of spatial intersection you chose will depend on the type of analysis you are doing. This concept is explained more deeply in the text book and in lecture.

*Exercise - Run the following code blocks, read the description of the operations and compare the outputs. Be prepared to discuss the relative utility of each operation based on its result.*

Note: you will likely get a warning about "attribute variables are assumed..." don't worry about it.

#### Spatial Clipping with `st_intersection()` on polygons

```{r}
clip <- 
  st_intersection(buffer, tracts16) %>%
  dplyr::select(TotalPop) %>%
  mutate(Selection_Type = "Clip")
```

#### Spatial intersection with `st_intersects()` on polygons

Simply, this type of spatial selection returns any polygon that overlaps or touches the buffer. We try to keep our code examples concise in this course, but here we show three different ways to achieve the same outcome.

1.  `selection1` uses `tracts16[buffer,]` uses the least code, but is
    often confusing unless you know the {sf} package well. In this
    usage, the intersecting rows of `tracts16` are selected for by
    indexing on `buffer`. This method is very fast to compute, but hard
    to understand at first.

2.  `selection2` is a longhand way to do the same. This uses
    `tracts16[st_intersects(tracts16, buffer) %>% lengths > 0, ]` to
    spell out all of the steps used in the first method. This can be
    read as: Select all rows of the rows from `tracts16` where
    `st_intersects()` returns a length of greater than zero. Important
    note that this uses `st_intersects()` and not `st_intersection()` as
    in the previous section. This approach is just as fast as approach
    #1, but a little easier to read.

3.  `selection3` takes a different approach and uses `st_join()` with the join type of "st_intersects". In this approach we need to follow the join with `filter(!is.na(Legend))` to remove the rows of `tracts16` where the polygons did not join. These are the polygons that did not intersect. This approach is the most readable and easiest to understand, but can be slower on large amounts of data.

```{r spatialSelection}

# Do a spatial selection to see which tracts touch the buffer

# approach #1: sub-setting a spatial object with a spatial object using the '[' brackets.
selection1 <- 
  tracts16[buffer,] %>%
  dplyr::select(TotalPop) %>%
  mutate(Selection_Type = "Spatial Selection")

# approach #2: using `st_intersects` as a verbose way to do approach #1
selection2 <- tracts16[st_intersects(tracts16, buffer) %>% lengths > 0, ] %>%
  dplyr::select(TotalPop) %>%
  mutate(Selection_Type = "Spatial Selection")

# approach #3: use `st_join` to do a spatial join and remove the non-intersecting polygons
selection3 <- tracts16 %>% 
  st_join(buffer, join = st_intersects) %>% 
  filter(!is.na(Legend)) %>% 
  dplyr::select(TotalPop) %>%
  mutate(Selection_Type = "Spatial Intersects")
```

#### Spatial intersection with with `st_centroid()` on polygon centroids

Finally, we are using centroids, or the point at the center of a polygon, to select whether a polygon from `tracts16` intersects the `buffer`. The steps here are:

1.  convert tract polygons to centroids and to the spatial intersect to buffers using the `selection1` approach from above. The resulting geometry type is now a point and not a polygon

2.  drop the geometry so that is becomes a regular non-spatial data frame

3.  join the original `tracts16` data frame using the `GEOID` as the join field. Note that only the `GEOID` column is selected, however the polygon spatial features are also joined.

4.  `st_sf()` is a function to turn it back to an {sf} spatial object

5.  finally the `TotalPop` column is selected and a new field of `Selection_Type` is added

Note: you will get a warning about "st_centroid assumes attributes... don't worry about it.

```{r}
selectCentroids <-
  st_centroid(tracts16)[buffer,] %>%
  st_drop_geometry() %>%
  left_join(., dplyr::select(tracts16, GEOID), by = "GEOID") %>%
  st_sf() %>%
  dplyr::select(TotalPop) %>%
  mutate(Selection_Type = "Select by Centroids")
```

Let's plot the results of each method to assess.

```{r}
intersections <- rbind(clip, selection1, selectCentroids)

ggplot() +
  geom_sf(data=intersections, aes(fill = TotalPop)) +
  geom_sf(data=septaStops, show.legend = "point") +
  scale_fill_viridis_c() +
  facet_wrap(~Selection_Type) + 
  mapTheme()
```

### Indicator Maps

We do our centroid joins as above, and then do a "disjoin" to get the ones that *don't* join, and add them all together. Do this operation and then examine it. What represents the joins/doesn't join dichotomy? Note that this contains a correct 2016-2020 inflation calculation. If you want to make an inflation adjustment for your homework, use [this inflation calculator from the BLS](https://www.bls.gov/data/inflation_calculator.htm).

```{r}
allTracts.group <- 
  rbind(
    st_centroid(allTracts)[buffer,] %>%
      st_drop_geometry() %>%
      left_join(allTracts) %>%
      st_sf() %>%
      mutate(TOD = "TOD"),
    st_centroid(allTracts)[buffer, op = st_disjoint] %>%
      st_drop_geometry() %>%
      left_join(allTracts) %>%
      st_sf() %>%
      mutate(TOD = "Non-TOD")) %>%
  mutate(MedRent.inf = ifelse(year == "2016", MedRent * 1.14, MedRent)) 

```
*Can you try to create the small multiple indicator maps seen in the text? The solutions are contained in "map_exercise.R"*

## Appendices - More Code For Completing Assignment 2

### TOD Indicator Tables

Your homework will require you to make indicator tables comparing TOD and non TOD areas over time. These code blocks create the tables you see in the book and in lecture.

```{r}
allTracts.Summary <- 
  st_drop_geometry(allTracts.group) %>%
  group_by(year, TOD) %>%
  summarize(Rent = mean(MedRent, na.rm = T),
            Population = mean(TotalPop, na.rm = T),
            Percent_White = mean(pctWhite, na.rm = T),
            Percent_Bach = mean(pctBachelors, na.rm = T),
            Percent_Poverty = mean(pctPoverty, na.rm = T))

kable(allTracts.Summary) %>%
  kable_styling() %>%
  footnote(general_title = "\n",
           general = "Table 2.2")
```

Let's make some comparisons and speculate about the willingness to pay and demographics in these areas 2016-2020 (see the 2000 data in the text too)

Notice how we pipe the kable() command here

```{r}
allTracts.Summary %>%
  unite(year.TOD, year, TOD, sep = ": ", remove = T) %>%
  gather(Variable, Value, -year.TOD) %>%
  mutate(Value = round(Value, 2)) %>%
  spread(year.TOD, Value) %>%
  kable() %>%
  kable_styling() %>%
  footnote(general_title = "\n",
           general = "Table 2.3")
```

### TOD Indicator Plots

For your homework, you will be required to create "small multiple" plots of these indicators. Let's use the "gather" command (look this one up - it's the opposite of `spread`) To go from wide to long data.

Why do we do this?? Notice we can "pipe" a ggplot call right into this operation!

```{r}
allTracts.Summary %>%
  gather(Variable, Value, -year, -TOD) %>%
  ggplot(aes(year, Value, fill = TOD)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~Variable, scales = "free", ncol=5) +
  scale_fill_manual(values = c("#bae4bc", "#0868ac")) +
  labs(title = "Indicator differences across time and space") +
  plotTheme() + 
  theme(legend.position="bottom")
```

#### Examining three submarkets

```{r}
centerCity <-
  st_intersection(
    st_buffer(filter(septaStops, Line == "El"), 2640) %>% st_union(),
    st_buffer(filter(septaStops, Line == "Broad_St"), 2640) %>% st_union()) %>%
  st_sf() %>%
  mutate(Submarket = "Center City")

el <-
  st_buffer(filter(septaStops, Line == "El"), 2640) %>% st_union() %>%
  st_sf() %>%
  st_difference(centerCity) %>%
  mutate(Submarket = "El")

broad.st <-
  st_buffer(filter(septaStops, Line == "Broad_St"), 2640) %>% st_union() %>%
  st_sf() %>%
  st_difference(centerCity) %>%
  mutate(Submarket = "Broad Street")

threeMarkets <- rbind(el, broad.st, centerCity)
```

You can then bind these buffers to tracts and map them or make small
multiple plots


```{r}
allTracts.threeMarkets <-
  st_join(st_centroid(allTracts), threeMarkets) %>%
  st_drop_geometry() %>%
  left_join(allTracts) %>%
  mutate(Submarket = replace_na(Submarket, "Non-TOD")) %>%
  st_sf() 
```

### Example of using `multipleRingBuffer()` function

There is a more comprehensive tutorial on the usage of this function (and the bugs you might encounter) in the Github folder!

```{r}
allTracts <- rbind(tracts20,tracts16)
septa_MRB <- multipleRingBuffer(st_union(septaStops), 47520, 2640)

allTracts.rings <-
  st_join(st_centroid(dplyr::select(allTracts, GEOID, year)),
          septa_MRB) %>%
  st_drop_geometry() %>%
  left_join(dplyr::select(allTracts, GEOID, MedRent, year), 
            by=c("GEOID"="GEOID", "year"="year")) %>%
  st_sf() %>%
  mutate(distance = distance / 5280) #convert to miles



  ggplot() +
    geom_sf(data=septa_MRB) +
    geom_sf(data=septaStops, size=1) +
    geom_sf(data=st_union(tracts20), fill=NA, size=1.2) +
    labs(title="Half mile buffers") +
    mapTheme()

```

